%% acknowledgments.tex

% From mitthesis package
% Version: 1.02, 2024/06/19
% Documentation: https://ctan.org/pkg/mitthesis

\chapter*{Acknowledgments}
\pdfbookmark[0]{Acknowledgments}{acknowledgments}

This thesis is the culmination of far too many sleepless nights, crash-outs, and hours spent staring at a computer screen over the past four years. But here we are: done. I do not know how to succinctly capture the immense gratitude I feel for everyone who has helped me along the way, but I hope this admittedly too-short acknowledgments section does it some justice.

First, thank you to Mami, Papi, Hali, Aby, G\"uelis, and Papa for always believing in me. You pushed me to succeed every step of the way, and I would not be here without you.

To all the friends and upperclassmen at MIT who let me rant at 2 a.m. and gave me space to be myself, thank you. To everyone in Simmons and East Campus, the Shenkman Hall woke suite, Lillian, everyone on 1E, DormCon, SIPB, WMBR, and TDC: you have helped me in more ways than I can count.

To Riccardo Fiorista and Gioele Zardini, thank you for introducing me to new fields and for being incredibly patient as I blundered my way through them. Thank you for giving me space to learn and grow as I discovered ideas I had never even heard of, and for coming to help when I finally admitted that I needed it. I could not have done this without the support you have given me throughout.

To Jim Aloisi, my thesis advisor: thank you for giving me feedback when I needed it most. To be honest, when I first took your class, I was a bit intimidated, but your coffee and kindness helped me through moments when I needed both. And to John Attanucci, who helped spearhead this project and took the lead when I did not know where to go: thank you for advocating for me and helping me along. I appreciate it more than you know.

To Cong Cong and John Akula, thank you for guiding me through my first research experience at MIT. You helped me lay the groundwork for this thesis and taught me to be confident in my work and advocate for myself. Whether working in Building 9 or driving to the Harvard Law Library, I will always carry those experiences with me.

To the staff and graduate students at MIT who supported me when I did not know who else to turn to: thank you. Some of you have seen me at my worst, and your care meant more than I can say. Thank you to Connor Perez, Amanda Putnam, Kat Howell, Kat Waxstein, Ana Lucena Navarro, Elaina Emery, Don Camelio, Helen Wang, Tasha Coppett, Elise Selinger, Ashley Kennedy, Maddie Garcia, Mich Lin, Rishi Kommalapati, and Briana Cuero.

To those at Volpe and WMATA who made me feel welcome even when the world felt like it was crashing down: thank you. At Volpe, thank you to Mark Mockett, Carmen Baskauf, Mac Lang, April Davidson, Hoai Tran, Nate Deshmukh Towery, Matt Cuddy, Michael York, Ian Berg, Scott Smith, Hannah Rakoff, and Rachael Sack. At WMATA, thank you to Doojin Han, Billy Gannom, and Jim Hankins.

To all the teachers before MIT who thought I could make it and pushed me to get here: we did it. Thank you to Ms. Bay, Ms. Camarena, Mr. D, and everyone at SciTech for supporting me and wishing me the best.

And finally, thank you to Skateboard, my cat of three years. You saw me at my best and worst, and you were one of the best parts of this hellish place.

% \begin{itemize}
%     \item Family (Mami, Papi, Hali, Aby, Guelis, Papa) for always believing in me and pushing me to succeed
%     \item Friends and upperclassmen at MIT for letting me rant at 2am and giving me space to be myself (Simmons ppl, gay agenda, 1E, Stickman, DormCon, SIPB, WMBR, TDC, Lillian)
%     \item Riccardo and Gioele for introducing me to new fields and being patient as I blundered my way through them
%     \item Jim and John for advising and giving me feedback when I needed it most
%     \item Cong Cong and John Akula for helping me through my first ever research experiences
%     \item Staff at MIT for always supporting me when I didn't know who to turn to (Connor, Amanda, Kat H, Kat W, Helen, Tasha, Elise, Ashley, Maddie Garcia, Mich, Rishi, Briana, Ana, and Elaina)
%     \item Those at Volpe and WMATA who always made me feel welcome even when it felt like the world was crashing down (Volpe 331/336: Mark, Carmen, Mac, April, Mikayla, Michael, Scott, Ian, Rachael) (WMATA: Doojin, Billy, Jim)
%     \item Teachers before MIT who thought I could make it (Ms. Bay, Ms. Camarena, Mr. D, everyone at SciTech)
%     \item Skateboard the cat
% \end{itemize}












