\chapter{Conclusions and Limitations}
\label{ch:conclusions}

\begin{table}[!htbp]
\centering
\caption{Limitations, effects on results, and recommended next steps.}
\label{tab:limitations}
\small
\begin{tabularx}{\textwidth}{p{0.20\textwidth} X p{0.25\textwidth}}
\toprule
\textbf{Limitation} & \textbf{Effect on Results} & \textbf{Next Step} \\
\midrule
Estimated travel times &
Route costs and feasibility may differ from real bus operations, especially during congestion and school arrival windows. &
Calibrate travel times with bus GPS traces. \\

Single demand snapshot &
Results may not capture absences, address changes, program changes, or midyear enrollment shifts. &
Test multiple student-assignment snapshots. \\

Incomplete operational constraints &
Routes may omit practical details such as driver preferences, depot procedures, dispatch flexibility, and special-case handling. &
Validate routes with district operators and encode missing constraints. \\

Limited uncertainty modeling &
Scenario results do not fully capture stochastic delays, disruptions, or day-to-day travel-time variability. &
Evaluate routes under sampled traffic and delay scenarios. \\

Case-study scale only &
Runtime and solution quality may differ for larger districts, finer networks, or more scenario samples. &
Benchmark on larger synthetic and real instances. \\

Indirect equity measures &
Optimization outputs may not fully capture reliability, walking burden, household constraints, or lived experience. &
Add stakeholder-defined equity metrics and thresholds. \\

No field deployment &
Generated routes have not been tested with drivers, families, or dispatchers in live service. &
Pilot or shadow-test selected routes before adoption. \\
\bottomrule
\end{tabularx}
\end{table}

\autoref{tab:limitations} summarizes the principal limitations of this study and recommended next steps. With those caveats, the Framingham case study demonstrates that the \gls{bra} implementation can expose operational tradeoffs that are difficult to see from the existing route plan alone. For the currently assigned bus riders, BRA serves the same 4,780 students using 48 buses rather than 51, a reduction of 3 buses, or approximately 5.9\%. At the per-bus daily lease cost discussed in \hyperref[ap:stakeholder]{Appendix~\ref*{ap:stakeholder}} (low hundreds of dollars), this corresponds to a meaningful annual lease-cost reduction before accounting for fuel and labor. At the same time, estimated mean travel time falls from 43.08 minutes to 30.52 minutes, and the maximum observed travel time falls from 133.38 minutes to 58.31 minutes. Under the modeled assumptions and conservative operational constraints, this suggests that the current routing pattern is improved upon by feasible alternatives that use fewer vehicles while also reducing student ride times---a more efficient and equitable system.

The all-students experiments show that expanding transportation access creates a different tradeoff. Serving students who live at least 1.5 miles from school is feasible under the modeled assumptions, though closer to the fleet limit. In \autoref{tab:travel_times_all}, the model uses 62 buses for the 1.5-mile eligibility case, leaving no students unassigned under the assumed constraints. When all students are considered, the model serves a larger total population but leaves 13 students unassigned, indicating that universal default service is not achievable without either additional resources, relaxed service standards, revised stop assignment rules, or a different policy definition of eligibility.

These results also illustrate the value of the Pareto-front framing. Rather than producing a single ``best'' routing plan, the co-design formulation can show the frontier between competing objectives. This is useful for planning because the preferred solution may depend on institutional priorities: one configuration may minimize buses, another maximizes student coverage, and another could reduce long rides for vulnerable groups. By placing the current \gls{fps} routing plan on the same frontier as generated alternatives, the framework allows planners to distinguish between tradeoffs that are structurally unavoidable and inefficiencies that can be improved through redesign.

Equity considerations are central to this interpretation. The preliminary results show that students from non-English-speaking households, non-car-owning households, and students with disabilities all experience lower estimated mean travel times under \gls{bra} than under the current routing baseline. This suggests that efficiency improvements do not come at the expense of vulnerable groups. However, the current experiments evaluate equity primarily after routing, rather than embedding equity constraints directly in the optimization objective. Future versions should incorporate explicit constraints or penalties for unequal service outcomes, such as subgroup-specific maximum ride times, minimum coverage rates, or limits on the spatial concentration of unassigned students. Additionally, inspired by the work of \textcite{vanweeEvaluatingTransportEquity2021}, promising outcomes could be gained by identifying ways of indexing inequity levels and incorporating them into \gls{bra}.

The results should therefore be interpreted as planning evidence rather than deployment-ready schedules. The current model assumes a constant travel speed, does not include intersection delay or time-varying congestion, and lacks complete student-level linkage between home addresses, current stops, bus assignments, disability status, and wheelchair-accessibility requirements. These limitations affect the precision of individual travel-time estimates and may change which route chains are feasible in practice. Before implementation, the model should be recalibrated using district-validated stop assignments, complete accessibility data, and travel-time estimates derived from \gls{avl} data or a time-dependent routing engine.

Overall, the case study supports the central claim of this work: a co-design-based routing framework can help districts reason about bus network redesign under multiple objectives. BRA provides a practical backend for generating feasible routing alternatives, while the Pareto-front representation clarifies how operational efficiency, coverage expansion, and equity goals interact. For \gls{fps}, the preliminary results indicate that meaningful improvements may be possible, particularly for currently assigned riders, but that broader opt-out service requires explicit policy choices about eligibility thresholds, fleet deployment, maximum ride times, and equity guarantees.

Taken together, this thesis contributes a compositional framework (grounded in the monotone co-design theory of \textcite{censi2016Mathematical}) for analyzing school bus routing as a co-design problem rather than a standalone vehicle-routing task. The \gls{mdpi} decomposition (fleet, monitor, driver, fuel, maintenance, and routing service components composed via shared functionality/resource interfaces, as in \autoref{fig:codesign}) is what enables planners to vary one component at a time and read off the resulting Pareto front, rather than re-solving an entire monolithic \gls{milp} for each scenario. By adapting the \gls{bird} backend into the \gls{bra}, the work connects district resources, service requirements, accessibility needs, emissions, and cost within a single tradeoff-oriented optimization pipeline. The Framingham case study demonstrates that this framework can translate real district data into feasible alternative routing designs and Pareto fronts that make the consequences of different planning priorities explicit. For planners, this enables a more transparent form of analysis: instead of asking only for the lowest-cost routes, districts can evaluate how operational, environmental, and equity objectives interact before selecting routes for further testing or implementation.

\section{Future Work}

Key next steps include (1) calibrating travel times with \gls{fps}'s bus \gls{gps} traces, (2) embedding equity directly in the optimization objective (e.g., concentration index or Rawlsian leximax following \textcite{paviaDesigningEquitableTransit2023}), (3) extending the case study beyond Framingham, and (4) piloting selected routes with district operators.